Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia

### 2025-10-11 _manuscript/manuscript.qmd
Extensive answer to R2/3 about missingness added to the manuscript.

---
lang: en
title: "Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia"
shorttitle: "Parkinson's Disease Dementia Level I Criteria"
author:
  - name: Martina Mana
    corresponding: false
    orcid: "0009-0007-4665-3946"
    role:
      - Conceptualization
      - Data curation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Josef Mana
    corresponding: false
    orcid: "0000-0002-7817-3978" 
    email: "josef.mana@lf1.cuni.cz"
    role:
      - Conceptualization
      - Data curation
      - Investigation
      - Formal analysis
      - Software
      - Methodology
      - Project administration
      - Validation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Tereza Uhrova
    corresponding: false
    email: "tereza.uhrova@vfn.cz"
    role:
      - Investigation
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Robert Jech
    corresponding: false
    orcid: "0000-0002-9732-8947"
    email: "jech@cesnet.cz"
    role:
      - Funding acquisition
      - Resources
      - Writing - review & editing
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Ondrej Bezdicek
    corresponding: true
    orcid: "0000-0002-5108-0181"
    email: "ondrej.bezdicek@gmail.com"
    role:
      - Investigation
      - Data curation
      - Funding acquisition
      - Conceptualization
      - Project administration
      - Supervision
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
abstract: ""
keywords:
  - Parkinson’s disease
  - Parkinson's disease dementia
  - level I criteria
format:
 # apaquarto-pdf:
   # documentmode: man
   # keep-tex: true
 apaquarto-docx: 
    toc: false
author-note:
  disclosures:
    financial-support: "Supported by The project National Institute for Neurological Research (Programme EXCELES, ID Project No. LX22NPO5107) - Funded by the European Union – Next Generation EU; Charles University: Cooperatio Program in Neuroscience; General University Hospital in Prague project MH CZ-DRO-VFN64165."
bibliography: bibliography.bib
floatsintext: false
numbered-lines: false
suppress-title-page: false
warning: false
echo: false
---

@@

data <- tar_read(raw_data, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$algorithms
vars <- tar_read(variables, store = upstore)
```

@@

```{r}
#| label: "crit-check"

critnum <- sapply(c("mmse", "moca", "smoca", "lvlII"), function(i) {
  subset(crit, group == i) |>
    nrow()
})

continue <- !any(critnum == 0)
stopifnot("Some algorithms are missing" = continue)
```

For each of these diagnostic approaches, we applied two operationalizations of deficits in Instrumental Activities of Daily Living (IADL). First, we utilized FAQ item 9, which approximates the pill questionnaire from the original criteria [@dubois2007] employing a cut-off score of 2 points or higher. Second, we applied the entire Functional Activities Questionnaire (FAQ) as suggested in the call for change [@kulisevsky2024], employing a cut-off score of 7 points or higher based on Czech normative data [@bezdicek2011]. These methodologies resulted in a total of `r sum(critnum)` algorithms, which were distributed across different diagnostic criteria: `r critnum["mmse"]` MMSE-based, `r critnum["moca"]` MoCA-based, `r critnum["smoca"]` sMoCA-based, and `r critnum["lvlII"]` based on the Level II battery (see @tbl-crits and Appendix @tbl-algos for the exact specification of each algorithm).

@@

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column). PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algorithm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

@@

Data wrangling and visualizations were done in the *tidyverse* package [@wickham2019] and tables were formatted in the *gt* package [@iannone2024]. All analyses were conducted within the R (version `r with(version, paste(major, minor, sep="."))`) software environment for statistical computing [@rsoft]. The software code supporting this article is available at <https://github.com/josefmana/DemCr1t.git>.[^1]

[^1]: **JM: Do not forget to make it public before submitted!**

@@

desc  <- tar_read(sample_description, store = upstore)
rates <- tar_read(rate_summaries, store = upstore)
algos <- tar_read(algorithms, store = upstore)
pred  <- tar_read(demographic_predictors, store = upstore)
conc  <- tar_read(concordance_statistics, store = upstore)
kappa <- tar_read(kappa_summmaries, store = upstore)
perc  <- rates$table$perc

@@

A total of `r sum(data$incl)` patients were included. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. **Descriptive statistics for neuropsychiatric symptoms indicated mild depressive and anxiety symptoms in our cohort, with the average BDI-II of `r bdi`, average STAI X1 of `r staix1`, and average STAI X2 of `r staix2`. However, according to the psychiatric assessment, none of the patients with probable PDD was suffering major depressive disorder, delirium or other neuropsychiatric abnormalities that would disqualify the diagnosis.** Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

Algorithm-wise rate of PDD estimates are presented in @tbl-rates. On average, estimated PDD rate was `r rates_overall["M"]`% (SD = `r rates_overall["SD"]`, Md = `r rates_overall["Md"]`, range `r rates_overall["minmax"]`). Notably, the estimates were substantially lower when FAQ item 9 was used as a criterion of IADL deficit (M = `r rates_FAQ9["M"]`% SD = `r rates_FAQ9["SD"]`, Md = `r rates_FAQ9["Md"]`, range `r rates_FAQ9["minmax"]`) compared to using total FAQ score criterion (M = `r rates_FAQtot["M"]`% SD = `r rates_FAQtot["SD"]`, Md = `r rates_FAQtot["Md"]`, range `r rates_FAQtot["minmax"]`) as demonstrated also in @fig-rates. Neither age nor sex or their interaction (*p*s $\geq$ `r min_p`) reliably predicted probable PDD classification across algorithms (see @fig-data and @fig-pars).

@@

@@

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **Numerical results are available at ...** [^2]. Generally, algorithms that employed the same operationalization of IADL deficit showed substantial pairwise concordance, however, algorithms that operationalized IADL deficit differently did not. Whereas among algorithms with identical IADL deficit operationalization, the agreement judged by Cohen's $\kappa$ was moderately high (operationalization by FAQ total score: $\kappa$ = `r kappa$sums["faq_tot"]`; operationalization by FAQ item 9: $\kappa$ = `r kappa$sums["faq_9"]`), among algorithms that differ in IADL deficit operationalization but are otherwise identical it was low: $\kappa$ = `r kappa$sums["iadl"]`.

[^2]: **JM: Should provide an excel sheet or html table the readers could go through by themselves.**

\[Insert Figure 2 here\]

@@

@@

??Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. Future studies would benefit from focusing on these aspects of PDD.??^[**JM: Why was the previous paragraph added instead of this one? This looks perfectly functional. The same goes largely for the IADL discussion, the original was better (in terms of resources and logical coherence) at most places.**]

>>>>

---
lang: en
title: "Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia"
shorttitle: "Parkinson's Disease Dementia Level I Criteria"
author:
  - name: Martina Mana
    corresponding: false
    orcid: "0009-0007-4665-3946"
    role:
      - Conceptualization
      - Data curation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Josef Mana
    corresponding: false
    orcid: "0000-0002-7817-3978" 
    email: "josef.mana@lf1.cuni.cz"
    role:
      - Conceptualization
      - Data curation
      - Investigation
      - Formal analysis
      - Software
      - Methodology
      - Project administration
      - Validation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Tereza Uhrova
    corresponding: false
    email: "tereza.uhrova@vfn.cz"
    role:
      - Investigation
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Robert Jech
    corresponding: false
    orcid: "0000-0002-9732-8947"
    email: "jech@cesnet.cz"
    role:
      - Funding acquisition
      - Resources
      - Writing - review & editing
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Ondrej Bezdicek
    corresponding: true
    orcid: "0000-0002-5108-0181"
    email: "ondrej.bezdicek@gmail.com"
    role:
      - Investigation
      - Data curation
      - Funding acquisition
      - Conceptualization
      - Project administration
      - Supervision
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
abstract: ""
keywords:
  - Parkinson’s disease
  - Parkinson's disease dementia
  - level I criteria
format:
 apaquarto-pdf:
    documentmode: man
    keep-tex: true
 apaquarto-docx: 
    toc: false
author-note:
  disclosures:
    financial-support: "Supported by The project National Institute for Neurological Research (Programme EXCELES, ID Project No. LX22NPO5107) - Funded by the European Union – Next Generation EU; Charles University: Cooperatio Program in Neuroscience; General University Hospital in Prague project MH CZ-DRO-VFN64165."
bibliography: bibliography.bib
floatsintext: false
numbered-lines: false
suppress-title-page: false
warning: false
echo: false
---

@@

data <- tar_read(raw_data, store = upstore)
vars <- tar_read(variables, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$algorithms
rates <- tar_read(rate_summaries, store = upstore)

pres <- tar_read(pdd_data, store = upstore)$PDD |>
  mutate(present = if_else(is.na(PDD), 0, 1)) |>
  select(id, type, present) |>
  pivot_wider(names_from = type, values_from = present) |>
  column_to_rownames("id")
```

@@

```{r}
#| label: "algo-check"

algonum <- sapply(c("mmse", "moca", "smoca", "lvlII"), function(i) {
  subset(crit, group == i) |>
    nrow()
})

continue <- !any(algonum == 0)
stopifnot("Some algorithms are missing" = continue)
```

For each of these diagnostic approaches, we applied two operationalizations of deficits in Instrumental Activities of Daily Living (IADL). First, we utilized FAQ item 9, which approximates the pill questionnaire from the original criteria [@dubois2007] employing a cut-off score of 2 points or higher. Second, we applied the entire Functional Activities Questionnaire (FAQ) as suggested in the call for change [@kulisevsky2024], employing a cut-off score of 7 points or higher based on Czech normative data [@bezdicek2011]. These methodologies resulted in a total of `r sum(algonum)` algorithms, which were distributed across different diagnostic criteria: `r algonum["mmse"]` MMSE-based, `r algonum["moca"]` MoCA-based, `r algonum["smoca"]` sMoCA-based, and `r algonum["lvlII"]` based on the Level II battery (see @fig-tree, @tbl-crits and Appendix @tbl-algos for the exact specification of each algorithm).

@@

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(algonum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column). PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algorithm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

@@

**Missing data were handled by complete cases analysis and pairwise complete cases analysis. In other words, each univariate and analysis pairwise comparison used all available data.** Data wrangling and visualizations were done in the *tidyverse* package [@wickham2019] and tables were formatted in the *gt* package [@iannone2024]. All analyses were conducted within the R (version `r with(version, paste(major, minor, sep="."))`) software environment for statistical computing [@rsoft]. The software code supporting this article is available at <https://github.com/josefmana/demcrit.git>.

@@

desc  <- tar_read(sample_description, store = upstore)
algos <- tar_read(algorithms, store = upstore)
pred  <- tar_read(demographic_predictors, store = upstore)
conc  <- tar_read(concordance_statistics, store = upstore)
kappa <- tar_read(kappa_summmaries, store = upstore)
perc  <- rates$table$perc

# Total numbers:
n_total <- sum(data$incl)
id_excl <- names(which(rowSums(pres) == 0))
n_excl <- length(id_excl)
n_incl <- n_total - n_excl
if (n_excl == 1) {
  excl_txt <- paste0("one patient was excluded due to missing neuropsychological data, resulting in a final sample of ", n_incl, " patients")
} else if (n_excl > 1) {
  excl_txt <- paste0(english::english(n_excl), " patients were excluded due to missing neuropsychological data, resulting in a final sample of ", n_incl, " patients")
} else {
  excl_txt <- "no patients was excluded."
}

@@

**Total of** `r n_total` patients **were considered for the study, out of which `r excl_txt`**. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. **Descriptive statistics for neuropsychiatric symptoms indicated mild depressive and anxiety symptoms in our cohort, with the average BDI-II of `r bdi`, average STAI X1 of `r staix1`, and average STAI X2 of `r staix2`. However, according to the psychiatric assessment, none of the patients with probable PDD was suffering from the major depressive disorder, delirium or other neuropsychiatric abnormalities that would disqualify the diagnosis.** Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

```{r}
#| label: "observations-univariate"

obs <- rates$table|>
  count(N) |>
  mutate(perc_missing = glue::glue("{round(100 * (n_incl-N) / n_incl, 1)}%"))
```

Algorithm-wise rate of PDD estimates **corresponding numbers of patients with available data** are presented in @tbl-rates. **In total, there were `r obs[1, "n"]` algorithms based on `r obs[1, "N"]` complete observations (`r obs[1, "perc_missing"]` missing). One algorithm used `r obs[2, "N"]`, `r obs[3, "N"]`, and `r obs[4, "N"]` complete observations, respectively (`r obs[2, "perc_missing"]`, `r obs[3, "perc_missing"]` and `r obs[4, "perc_missing"]` missing). A further `r obs[5, "n"]` algorithms were based on `r obs[5, "N"]` complete observations (`r obs[5, "perc_missing"]` missing), and `r obs[6, "n"]` algorithms included the full sample of `r obs[6, "N"]` patients.**

On average, estimated PDD rate was `r rates_overall["M"]`% (SD = `r rates_overall["SD"]`, Md = `r rates_overall["Md"]`, range `r rates_overall["minmax"]`). Notably, the estimates were substantially lower when FAQ item 9 was used as a criterion of IADL deficit (M = `r rates_FAQ9["M"]`% SD = `r rates_FAQ9["SD"]`, Md = `r rates_FAQ9["Md"]`, range `r rates_FAQ9["minmax"]`) compared to using total FAQ score criterion (M = `r rates_FAQtot["M"]`% SD = `r rates_FAQtot["SD"]`, Md = `r rates_FAQtot["Md"]`, range `r rates_FAQtot["minmax"]`) as demonstrated in @fig-tree **(see also @fig-rates for per-operationalisation distribution of PDD rate estimates)**. Neither age nor sex or their interaction (*p*s $\geq$ `r min_p`) reliably predicted probable PDD classification across algorithms (see @fig-data and @fig-pars).

@@

```{r}
#| label: "fig-tree"
#| fig-cap: "A dendrogram representing algorithms creation."

make_dendrogram(rates$table)
```

@@

```{r}
#| label: "observations-bivariate"

dims <- glue::glue("{dim(conc$table)[1]} rows x {dim(conc$table)[2]} columns")

obs_pairs <- conc$table |>
  count(N) |>
  drop_na() |> # NA for cases where reference == predictor
  mutate(perc = 100 * n / sum(n))

min_miss <- glue::glue("{round(100 * (n_incl - obs_pairs[1, 'N']) / n_incl, 2)}%")

K <- 200
perc_above_k <- obs_pairs |>
  filter(N >= K) |>
  pull(perc) |>
  sum() |>
  do_summary(2)
```

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **The number of complete pairwise cases ranged from `r obs_pairs[1, "N"]` (`r min_miss` missing) to `r obs_pairs[nrow(obs_pairs), "N"]` (full sample). Most comparisons (`r perc_above_k`%) were based off of `r K` or more observations.**[^1]

[^1]: **Due to the large number of entries (`r dims` representing pairwise comparisons and metrics of interest respectively), the table with numerical results is not presented here or in the Appendix. Instead, we share the table share as data in the accompanying R package available at <https://github.com/josefmana/demcrit.git>. To obtain the table in format not dependent on R, follow the tutorial at <https://josefmana.github.io/demcrit/articles/concordance.html>**

Generally, algorithms that employed the same operationalization of IADL deficit showed substantial pairwise concordance, however, algorithms that operationalized IADL deficit differently did not. Whereas among algorithms with identical IADL deficit operationalization, the agreement judged by Cohen's $\kappa$ was moderately high (operationalization by FAQ total score: $\kappa$ = `r kappa$sums["faq_tot"]`; operationalization by FAQ item 9: $\kappa$ = `r kappa$sums["faq_9"]`), among algorithms that differ in IADL deficit operationalization but are otherwise identical it was low: $\kappa$ = `r kappa$sums["iadl"]`.

\[Insert Figure 3 here\]

@@

**Due to the retrospective nature of the study, some patients lacked one or more key measures required for the diagnosis of probable PDD by certain algorithms. Missing data were handled using the pairwise complete cases method. The main advantages of this approach are its straightforward implementation and preservation of statistical power. However, it may introduce bias, particularly when data are not missing completely at random (MCAR) and when causal inference is the goal [@little2019].**

**Modern alternatives, such as multiple imputation techniques, have been shown to produce less biased estimates than case deletion strategies in analyses based on confusion matrices [@karakaya2014], whereas the evidence supporting their superiority in the estimation of Cohen’s $\kappa$ remains limited [@deraadt2019]. These modern approaches also require careful specification of the imputation model and the underlying causal mechanism of missingness to ensure appropriate covariate selection [@long2011; @bianco2023]. For the sake of parsimony, we did not employ advanced missing-data techniques but instead explicitly described the observed missingness patterns. Consequently, our findings should be regarded as exploratory and primarily serve as a basis for hypothesis generation in future, more controlled studies.**

**Another possible avenue for analysing the present data would be to exploit their hierarchical structure by fitting a multilevel model with algorithms nested within patients (see @fig-tree). Such an approach could leverage partial pooling [@gelman2012] and might approximate modern item response theory frameworks [@burkner2021], where PDD is conceptualised as a latent trait and the individual diagnostic algorithms act as items measuring it (). If feasible, this psychometric perspective may represent an exciting new direction for conceptualising complex clinical phenomena such as PDD [cf. @miller2012, @kiselica2021].**

@@

??Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. Future studies would benefit from focusing on these aspects of PDD.??[^2]

[^2]: **JM: Why was the previous paragraph added instead of this one? This looks perfectly functional. The same goes largely for the IADL discussion, the original was better (in terms of resources and logical coherence) at most places.**
